---
title: "Data Manipulation with dplyr"
author: "Vahan Aslanyan"
format: 
  revealjs:
    theme: dark
---

## Data Wrangling and Manipulation

- Makes data accessible to users

- Creates final dataset(s) for projects (analysis,report,etc)

- Create new variables or summaries

- Rename the variables or observations

## Data Frames

- Data frames are the most common structures used for analyses

- Identifiable columns (with preferably non-empty headers) containing variables of interest

- Rows containing observed values of variables of interest

- Mixed formats allowed (strings, dates, numeric values); matrices can be transformed into dataframe

## Example

```{r}
#| echo: true
#| output-location: fragment
# to install: install.packages("nycflights13")
# to install: install.packages("tidyverse")
library(nycflights13)
library("dplyr")
top_n(flights,10) #from nycflight13 package
```


## Data manipulation--Reasons

- Original data is not ready to use

- Maybe you are interested in a subset of variables

- Maybe you are only interested in specific observations

- Maybe you need to create new variables (morning arrivals vs evening arrivals)

- Maybe you need to fix the encoding of the variables (arr_time is an integer now, we need to have it as time)

## Data manipulation--Motivation

- Certain effort goes into data manipulation

- 80/20 rule: 80% of time is spent cleaning up data, 20%--doing analyses or generating insights

- We want a clean readable code

- We want our steps to be intuitive

## Dplyr package

- "Grammar of data manipulation"

- Provides a consistent set of verbs that help solve the most common data manipulation challenges

- <https://dplyr.tidyverse.org/>

- Cheat sheet: <https://github.com/rstudio/cheatsheets/blob/master/data-transformation.pdf>

## Dplyr basics
- Pick observations by their values (filter())

- Reorder the rows (arrange())

- Pick variables by their names (select())

- Create new variables with functions of existing variables (mutate())

- Collapse many values down to a single summary (summarise())

These can all be used in conjunction with group_by() which changes the scope of each function from operating on the entire dataset to operating on it group-by-group.


## Filter rows with filter()

filter() allows to subset observations based on their values. For example we can select all flights on January first with:

```{r}
#| echo: true
#| code-line-numbers: "|1|2"
#| output-location: fragment
flights%>%
  filter(month==1,day==1)->flight_Jan01
top_n(flight_Jan01,10)  
```


## Comparisons
R useses the standard suite for comparisons

- `>` greater than
-  `>=` greater than or equal to

- < less than

- <= less than or equal to

- == equal

- != not equal



## Exercise: Early bird

A traveler always has meetings at 10 am so she has to arrive before 8:30 to be there on time. She is planning a trip in May and wants to arrive in STL before then. Create a dataframe flights_830 that contains all May flights that arrive at STL before 8:30

## Exercise: Early bird
```{r}
#| echo: true
#| code-line-numbers: "|1|2"
#| output-location: fragment
flights%>%
  filter(month==5, arr_time<=830,dest=="STL")->flights_830
flights_830  
```


## Logical operators

- & and

- `|` or

- ! not

Now lets get all flights in November and December:


```{r}
#| echo: true
#| output-location: fragment
filter(flights, month == 11 | month == 12)
```

## Logical operators
```{r}
#| echo: true

#| output-location: fragment

flights%>%filter(month==11| month ==12)
```

## Logical operators

```{r}
#| echo: true
#| output-location: fragment

filter(flights, month %in% c(11, 12))
```

## Logical operators

If we wanted to find flights that weren’t delayed (on arrival or departure) by more than two hours, you could use either of the following two filters:
```{r}
#| echo: true
#| output-location: fragment
filter(flights, !(arr_delay > 120 | dep_delay > 120))
```

## Logical operators
```{r}
#| echo: true
#| output-location: fragment
filter(flights, arr_delay <= 120, dep_delay <= 120)
```


## Missing values

Missing values `NA`s make comparison tricky. If we want to determine if a value is missing, use is.na():

```{r}
#| echo: true
#| output-location: fragment
df <- tibble(x = c(1, NA, 3))
df%>%filter(is.na(x) | x > 1)
```


## Arrange rows with arrange()

- arrange() works similarly to filter() except that instead of selecting rows, it changes their order.
- It takes a data frame and a set of column names (or more complicated expressions) to order by. 
- If you provide more than one column name, each additional column will be used to break ties in the values of preceding columns.

```{r}
#| echo: true
#| output-location: fragment

flights%>%arrange(year, month, day)
```

## Arrange rows with arrange()-Continued

- Use desc() to re-order by a column in descending order:

```{r}
#| echo: true
#| output-location: fragment

flights%>%arrange(desc(dep_delay))
```

## Arrange rows with arrange()-Continued

- Missing values are always sorted at the end:

```{r}
#| echo: true
#| output-location: fragment

df <- tibble(x = c(5, 2, NA))
arrange(df, x)
```


## Select columns with select()

- Narrow in on the variables we are interested in

- Select() allows to rapidly zoom in on a useful subset using operations based on the names of the variables


```{r}
#| echo: true
#| output-location: fragment
# Select columns by name
select(flights, year,month, day)

```

## Select columns with select()

```{r}
#| echo: true
#| output-location: fragment
# Select all columns between year and day (inclusive)
select(flights, year:day)
```

## Select columns with select()

```{r}
#| echo: true
#| output-location: fragment
# Select all columns except those from year to day (inclusive)
select(flights, -(year:day))

```

## Select columns with select()

- There are a number of helper functions we can use within select():

- starts_with("abc"): matches names that begin with "abc"

- ends_with("xyz"): matches names that end with "xyz"

- contains("ijk"): matches names that contain "ijk"

- num_range("x", 1:3): matches x1, x2, and x3


## Select columns with select()

- everything() is useful for rearranging columns

```{r}
select(flights, time_hour, air_time, everything())
```


## Rename variables with rename()

```{r}
#| echo: true
#| output-location: fragment
rename(flights, tail_num = tailnum)
```



## Add new variable with mutate()

- Sometimes we need to create new columns that are function of existing columns

- mutate() always add new columns at the end of the dataset

```{r}
#| echo: true


flights_sml <- select(flights, 
  year:day, 
  ends_with("delay"), 
  distance, 
  air_time
)
```

## Add new variable with mutate()

```{r}
#| echo: true
#| code-line-numbers: "|1|2|3"
#| output-location: fragment
mutate(flights_sml,
  gain = dep_delay - arr_delay,
  speed = distance / air_time * 60
)


```

## Add new variable with mutate()
We can refer to columns that we have just created
```{r}
#| echo: true
#| code-line-numbers: "|1|2|3"
#| output-location: fragment
mutate(flights_sml,
  gain = dep_delay - arr_delay,
  hours = air_time / 60,
  gain_per_hour = gain / hours
)
```
## Add new variable with mutate()

We can use transmute() if we only want to keep the new variables

```{r}
#| echo: true
#| output-location: fragment
transmute(flights,
  gain = dep_delay - arr_delay,
  hours = air_time / 60,
  gain_per_hour = gain / hours
)
```

## Grouped summaries with summarise()

- summarise() collapses a data frame to a single row
```{r}
#| echo: true
#| output-location: fragment
summarise(flights, delay = mean(dep_delay, na.rm = TRUE))
```

## Grouped summaries with summarise()

- summarise() has limited usefulness unless we pair it with group_by()

- group_by()  changes the unit of analysis from the complete dataset to individual groups

```{r}
#| echo: true
#| code-line-numbers: "|1|2"
#| output-location: fragment
by_day <- group_by(flights, year, month, day)
summarise(by_day, delay = mean(dep_delay, na.rm = TRUE))
```


## Combining multiple operations with the pipe

We want to explore the relationship between the distance and average delay for each location

- group flights by destination

- summarise to compute distance, average delay, and number of flights

- filter to remove noisy points and Honolulu airports, which is almost twise as fat away as the next closest airport



## Combining multiple operations with the pipe

```{r}
#| echo: true
#| code-line-numbers: "|1|2|3|4|5|7"
#| output-location: fragment
by_dest <- group_by(flights, dest)
delay <- summarise(by_dest,
  count = n(),
  dist = mean(distance, na.rm = TRUE),
  delay = mean(arr_delay, na.rm = TRUE)
)
delay <- filter(delay, count > 20, dest != "HNL")
```

## Combining multiple operations with the pipe
```{r}
#| echo: true
#| code-line-numbers: "|1|2|3|4|5|6|7|8|9"
#| output-location: fragment
delays <- flights %>% 
  group_by(dest) %>% 
  summarise(
    count = n(),
    dist = mean(distance, na.rm = TRUE),
    delay = mean(arr_delay, na.rm = TRUE)
  ) %>% 
  filter(count > 20, dest != "HNL")
```

## Final exercise

Let's look at how the average performance of batters in baseball is related to the number of times they're at bat.

- Use Lahman package to compute the batting average of every major league baseball player

- create batting average variable `ba` (sum of hits(H)/sum of opportunities to his the ball (AB))

- create total opportunities to hit the ball variable `ab`(see above for definition)

- only keep batters with 100 and more opportunities


## Final exercise

```{r}
#| echo: true
#| code-line-numbers: "|1|2|3|4|5|6|7|8"
#| output-location: fragment
library(Lahman)
batting<-Lahman::Batting
batters <- batting %>% 
  group_by(playerID) %>% 
  summarise(
    ba = sum(H, na.rm = TRUE) / sum(AB, na.rm = TRUE),
    ab = sum(AB, na.rm = TRUE)
  )%>%filter(ab>100)
```

