---
title: "R Packages for Analysis"
date: "7/14/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Recommended R Packages

R has a thriving development community, and there are now over 12,000 packages on CRAN, the central repository for R packages. Many also exist on GitHub. Here are several we recommend for common analyses. Packages that live on GitHub need to be installed with the `devtools` package. Packages below can be installed from CRAN using `install.packages("package_name")` unless otherwise stated.

```{r eval=FALSE}
install.packages("devtools")
```

# Common regression models
- Linear Regression, no package needed: `lm()`
- GLMs like logistic and poisson, no package needed: `glm()`

# Longitudinal Analysis
- Mixed Models: `nlme`, `lme4`
- GEE: `gee`, `geepack`
- Survey Analysis: `survey`

# Survival Analysis
- `survival`

# Spatial Analysis
- Simple features that work well with `tidyverse` packages: `sf`
- Geocoding: the `geocode()` function in the `ggmap` package

# Pyschometrics and Latent Variable Modeling
- General pyschometric tools: `psycho`
- Item Response Theory: `ltm`
- Structural Equation Modeling: `lavaan`
- Latent Class Analysis: `poLCA`

# Bayesian Analysis
- Pre-compiled Bayesian models that look and act like other R functions: `rstanarm`
- Flexible Bayesian modeling: `brms`
- Visualization: `bayesplot`

# Genomic Analysis

# Color Palettes 
- Colorblind friendly palettes and tools to test palettes for color-blind friendliness: `colorblindr`

```{r eval=FALSE}
# requires development versions of cowplot and colorspace

devtools::install_github("wilkelab/cowplot")
install.packages("colorspace", repos = "http://R-Forge.R-project.org")

devtools::install_github("clauswilke/colorblindr")
```


- Palettes for reliable, scientific visualizations: `vidris` (and, in `ggplot2` 3.0.0, `scale_*_vidris_c()` and `scale_*_vidris_d()`), `scico`

```{r eval=FALSE}
devtools::install_github("thomasp85/scico")
```