---
title: "R Packages for Analysis"
date: "7/14/2018"
output: pdf_document
header-includes:
  - \usepackage{hyperref}
  - \hypersetup{colorlinks=true, allcolors=blue}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Recommended R Packages

R has a thriving development community, and there are now over 12,000 packages on CRAN, the central repository for R packages. Many also exist on GitHub. Here are several we recommend for common analyses. Packages that live on GitHub need to be installed with the `devtools` package. Packages below can be installed from CRAN using `install.packages("package_name")` unless otherwise stated.

```{r eval=FALSE}
install.packages("devtools")
```

# Common regression models

- Linear Regression, no package needed: `lm()`
- GLMs like logistic and Poisson, no package needed: `glm()`


# Longitudinal Analysis

- Mixed Models: [`nlme`](https://cran.r-project.org/package=nlme), [`lme4`](https://cran.r-project.org/package=lme4)
- GEE: [`gee`](https://cran.r-project.org/package=gee), [`geepack`](https://cran.r-project.org/package=geepack)
- Survey Analysis: [`survey`](https://cran.r-project.org/package=survey)


# Survival Analysis

- [`survival`](https://cran.r-project.org/web/packages/survival/index.html) (installed by default)


# Spatial Analysis

- [`cshapes`](https://cran.r-project.org/package=cshapes): The CShapes Dataset and Utilities
- [`spdep`](https://cran.r-project.org/package=spdep): Spatial Dependence: Weighting Schemes, Statistics and Models
- Simple features that work well with [`tidyverse`](https://cran.r-project.org/package=tidyverse) packages: [`sf`](https://cran.r-project.org/package=sf)
- Geocoding: the `geocode()` function in the [`ggmap`](https://cran.r-project.org/package=ggmap) package
- [`sphet`](https://cran.r-project.org/package=sphet): Estimation of Spatial Autoregressive Models with and without Heteroscedasticity
- [`spatialprobit`](https://cran.r-project.org/package=spatialprobit): Spatial Probit Models


# Pyschometrics and Latent Variable Modeling

- General psychometric tools: [`psych`](https://cran.r-project.org/package=psych)
- Item Response Theory: [`ltm`](https://cran.r-project.org/package=ltm), [`mirt`](https://cran.r-project.org/web/packages/mirt/index.html)
- Structural Equation Modeling: [`lavaan`](https://cran.r-project.org/package=lavaan)
- Latent Class Analysis: [`poLCA`](https://cran.r-project.org/package=poLCA)


# Bayesian Analysis

- Pre-compiled Bayesian models that look and act like other R functions: [`rstanarm`](https://cran.r-project.org/package=rstanarm)
- Flexible Bayesian modeling: [`brms`](https://cran.r-project.org/package=brms)
- Visualization: [`bayesplot`](https://cran.r-project.org/package=bayesplot)
- [`coda`](https://cran.r-project.org/package=coda): Output Analysis and Diagnostics for MCMC

# Econometrics

- [`AER`](https://cran.r-project.org/package=AER): Applied Econometrics with R

# Causal Inference
- [`sandwich`](https://cran.r-project.org/package=sandwich): Robust Covariance Matrix Estimators ([website](https://cran.r-project.org/package=sandwich))
- [`MatchIt`](https://cran.r-project.org/package=MatchIt): Nonparametric Preprocessing for Parametric Causal Inference
- [`Matching`](https://cran.r-project.org/package=Matching): Multivariate and Propensity Score Matching with Balance Optimization
- [`ipw`](https://cran.r-project.org/web/packages/ipw/ipw.pdf): Estimate Inverse Probability Weights

# Missing Data
- [`Amelia`](https://cran.r-project.org/package=Amelia): A Program for Missing Data
- [`mice`](https://cran.r-project.org/web/packages/mice/): Multivariate Imputation by Chained Equations
- [`naniar`](https://cran.r-project.org/web/packages/naniar/): Data Structures, Summaries, and Visualisations for Missing Data

# Social Networks

- [`igraph`](https://cran.r-project.org/package=igraph): Network Analysis and Visualization
- [`statnet`](https://cran.r-project.org/package=statnet): Software Tools for the Statistical Analysis of Network Data
- [`RSiena`](https://cran.r-project.org/package=RSiena): Siena - Simulation Investigation for Empirical Network Analysis


# High Performance Computing (and alike)

- [`Rcpp`](https://cran.r-project.org/package=Rcpp): Seamless R and C++ Integration
- [`RcppArmadillo`](https://cran.r-project.org/package=RcppArmadillo): 'Rcpp' Integration for the 'Armadillo' Templated Linear Algebra Library
- [`RcppParallel`](https://cran.r-project.org/package=RcppParallel): Parallel Programming Tools for 'Rcpp'
- [`biglm`](https://cran.r-project.org/package=biglm): bounded memory linear and generalized linear models


# Color Palettes 
- Colorblind friendly palettes and tools to test palettes for color-blind friendliness: [`colorblindr`](https://cran.r-project.org/package=colorblindr), [`viridis`](https://cran.r-project.org/package=viridis)

```{r eval=FALSE}
# requires development versions of cowplot and colorspace

devtools::install_github("wilkelab/cowplot")
install.packages("colorspace", repos = "http://R-Forge.R-project.org")

devtools::install_github("clauswilke/colorblindr")
```


- Palettes for reliable, scientific visualizations: [`vidris`](https://cran.r-project.org/package=vidris) (and, in [`ggplot2`](https://cran.r-project.org/package=ggplot2) 3.0.0, `scale_*_vidris_c()` and `scale_*_vidris_d()`), [`scico`](https://cran.r-project.org/package=scico)

```{r eval=FALSE}
devtools::install_github("thomasp85/scico")
```
