---
title: "R Packages for Analysis"
date: "7/14/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Recommended R Packages

R has a thriving development community, and there are now over 12,000 packages on CRAN, the central repository for R packages. Many also exist on GitHub. Here are several we recommend for common analyses. Packages that live on GitHub need to be installed with the `devtools` package. Packages below can be installed from CRAN using `install.packages("package_name")` unless otherwise stated.

```{r eval=FALSE}
install.packages("devtools")
```

# Common regression models

- Linear Regression, no package needed: `lm()`
- GLMs like logistic and Poisson, no package needed: `glm()`


# Longitudinal Analysis

- Mixed Models: `nlme`, `lme4`
- GEE: `gee`, `geepack`
- Survey Analysis: `survey`


# Survival Analysis

- `survival`


# Spatial Analysis

- cshapes: The CShapes Dataset and Utilities
- spdep: Spatial Dependence: Weighting Schemes, Statistics and Models
- Simple features that work well with `tidyverse` packages: `sf`
- Geocoding: the `geocode()` function in the `ggmap` package
- sphet: Estimation of Spatial Autoregressive Models with and without Heteroscedasticity
- spatialprobit: Spatial Probit Models


# Pyschometrics and Latent Variable Modeling

- General psychometric tools: `psycho`
- Item Response Theory: `ltm`
- Structural Equation Modeling: `lavaan`
- Latent Class Analysis: `poLCA`


# Bayesian Analysis

- Pre-compiled Bayesian models that look and act like other R functions: `rstanarm`
- Flexible Bayesian modeling: `brms`
- Visualization: `bayesplot`
- coda: Output Analysis and Diagnostics for MCMC ([website](https://cran.r-project.org/package=coda))


# Genomic Analysis


# Econometrics

- AER: Applied Econometris with R ([website](https://cran.r-project.org/package=AER))
- sandwich: Robust Covariance Matrix Estimators ([website](https://cran.r-project.org/package=sandwich))
- MatchIt: Nonparametric Preprocessing for Parametric Causal Inference
- Amelia: A Program for Missing Data
- Matching: Multivariate and Propensity Score Matching with Balance Optimization


# Social Networks

- igraph: Network Analysis and Visualization
- statnet: Software Tools for the Statistical Analysis of Network Data
- RSiena: Siena - Simulation Investigation for Empirical Network Analysis


# High Performance Computing (and alike)

- Rcpp: Seamless R and C++ Integration
- RcppArmadillo: 'Rcpp' Integration for the 'Armadillo' Templated Linear Algebra Library
- RcppParallel: Parallel Programming Tools for 'Rcpp'
- biglm: bounded memory linear and generalized linear models


# Color Palettes 
- Colorblind friendly palettes and tools to test palettes for color-blind friendliness: `colorblindr`, `viridis`

```{r eval=FALSE}
# requires development versions of cowplot and colorspace

devtools::install_github("wilkelab/cowplot")
install.packages("colorspace", repos = "http://R-Forge.R-project.org")

devtools::install_github("clauswilke/colorblindr")
```


- Palettes for reliable, scientific visualizations: `vidris` (and, in `ggplot2` 3.0.0, `scale_*_vidris_c()` and `scale_*_vidris_d()`), `scico`

```{r eval=FALSE}
devtools::install_github("thomasp85/scico")
```
