---
title: 'R Bootcamp day 3<br>Reproducible Research<br><img src="../../fig/trojan-rlogo.svg" style="width:250px;">'
author: <a href="https://ggvy.cl"><b>George G. Vega Yon</b></a><br>vegayon@usc.edu<br><br>University of Southern California<br>Department of Preventive Medicine
date: August 15th, 2018
output: 
  slidy_presentation:
    footer: R Bootcamp (Day 2)
    highlight: pygments
    font_adjustment: -1
---

## Reproducible Research

> "The term reproducible research refers to the idea that the ultimate product
> of academic research is the paper along with the __full computational environment
> used to produce the results in the paper such as the code, data, etc.__ that
> can be used to reproduce the results and __create new work based on the research__"

\raggedleft 

[Wiki](https://en.wikipedia.org/wiki/Reproducibility#Reproducible_research)

## Reproducible Research (cont. II) {.smaller}

-   A major new issue in sciences (overall):
    * Accessible Reproducible Research ([Mesirov, __Science__ 2010](http://science.sciencemag.org/content/327/5964/415))
    * Again, and Again, and Again, ... ([Jasny et al., __Science__ 2011](http://science.sciencemag.org/content/334/6060/1225))
    * Challenges in Irreproducible Research ([__nature__ topic](http://www.nature.com/news/reproducibility-1.17552))
    
## How to 'Reproducible Research'

- What you can do:
    * Provide __raw__ data (raw, i.e. before "cleaning it"),
    * Provide source code (what ever programming environment you are using) for reproducing:
  _cleaned data_, models, tables, figures, etc.
    * Hopefully have a neat way of coding your programs: Inline Comments, Indentation of control-flow statements (for, while, case, switch, ifelse, etc.)
- What else
    * Try using version control software (such as git) to make your research "opensource"
    * Avoid using proprietary software (hopefully always)


- R
    * Try using [knitr](https://cran.r-project.org/web/packages/knitr/index.html) and [Rmarkdown](https://cran.r-project.org/web/packages/rmarkdown/index.html)
    * [texreg](https://cran.r-project.org/web/packages/texreg/index.html) for fancy regression tables.
    * Checkout `?grDevices::Devices`.
    * More resources at CRAN task View [Reproducible Research](https://cran.r-project.org/web/views/ReproducibleResearch.html)

# Research pipeline: Regression tables with texreg

*   The R package [**texreg**](https://cran.r-project.org/package=texreg){target="_blank"}

```{r multiple-regressions, warning=FALSE, message=FALSE}
library(texreg)
data(diamonds, package="ggplot2")

# Model 1
model1 <- lm(price ~ carat, data = diamonds)
model2 <- lm(price ~ carat + depth, data = diamonds)
model3 <- lm(price ~ carat + table, data = diamonds)
model4 <- lm(price ~ carat + depth + table, data = diamonds)

# Let's put it all in a list to handle it together
models <- list(model1, model2, model3, model3)
```

---

```{r table-html, results='asis'}
htmlreg(models, doctype=FALSE)
```

It also has the functions `texreg`, for LaTeX tables, and `screenreg`, for plaintext output

# Research pipeline: Regression tables with xtable