---
title: 'R Bootcamp day 2<br>Statistical Computing Track<br><img src="../fig/trojan-rlogo.svg" style="width:250px;">'
author: <a href="https://ggvy.cl"><b>George G. Vega Yon</b></a><br>vegayon@usc.edu<br><br>University of Southern California<br>Department of Preventive Medicine
date: August 20th, 2019
output: 
  slidy_presentation:
    footer: R Bootcamp (Day 2)
    highlight: pygments
    font_adjustment: -1
---

# A recap from Day 1

1.  R has 6 atomic types...
    
2.  R Core has several functions for simulating and computing statistical
    distributions. The names start with...
    
3.  Overall, we should use (blank) instead of for-loops.

4.  We should set `set.seed` (blank).
    
  
# Beyond the vector

A very important data-type... matrices. You tell me about them

*  What are matrices in R?

*  How are matrices stored in R?

*  How many fundamental matrix types live in core R?

# Beyond the vector (cont'd)

"Fundamental" types of matrices:

1.  **Dense ("regular") matrix** Fully supported by base R

2.  **Sparse matrix** Only store the non-zero values. Can be row or column
    major... the Compress colum-major class `dgCMatrix` in the Matrix package,
    for example.
    
    Two main R packages: [Matrix](https://cran.r-project.org/package=Matrix) and
    [SparseM](https://cran.r-project.org/package=SparseM).
    
    **Matrix** is fully supported by
    [RcppArmadillo](https://cran.r-project.org/package=RcppArmadillo) and
    [RcppEigen](https://cran.r-project.org/package=RcppEigen) (will talk more
    about these on Wednesday!)
    
    If you work with spatial data... then these are your friends!
    
3.  **File-backed matrices** Usually dense, but doesn't fit in memory!
    One package for working with this: [bigmemory](https://cran.r-project.org/package=bigmemory).
    
    Also supported by RcppArmadillo and RcppEigen (I believe?).
   

# Matrix algebra 101

*   **Matrix**: A special class of array, are vectors with a `dim` attribute of length 2 (number of rows and number of columns).
    
    ```{r}
    mymat <- matrix(1:9, ncol=3) # A matrix with numbers 1 through 9 with 3 columns
    mymat
    str(mymat)
    ```
    
    R stores elements in column major-order. We can access matrices elements in the following way.
    
    ```{r, collapse=TRUE}
    mymat[1, 2] # The element in the (1, 2) location of the matrix
    mymat[4]    # The fourth element in the vector... which is (1, 2)
    ```
    
    Matrices can have row and column names too.
  
# Matrix algebra 101 (cont'd)

*   Some matrix fundamental operators, for matrices `A`, and `B` (both square) with the same dimensions (just for the examples):

    *   Element wise product, addition, substraction, and quotient: `A * B`, `A + B`, `A - B`, `A/B`
    *   Transpose `t(A)`.
    *   Inverse `solve(A)`.
    *   Column and row bind, `cbind(A, B)`, `rbind(A, B)`.
    *   Matrix product between both `A %*% B`.
    
*   Your favorite model can be simulated using linear algebra!
    
    ```{r eg-ols-sim}
    # Simulating an Linear Model
    set.seed(181)
    X <- cbind(1, rnorm(1000))
    y <- X %*% cbind(c(2.3, 5)) + rnorm(1000)
    
    # Estimating the model using the lm function
    coef(lm(y ~ -1 + X))
    
    # O a la linear algebra :)
    solve(t(X) %*% X) %*% t(X) %*% y
    ```


# Matrix problems:

*   Given the following matrices `A`, `B`, and C
    
    ```{r, R.options=list(digits=2)}
    set.seed(122)
    A <- matrix(rnorm(12), ncol=3)
    B <- matrix(rnorm(12), nrow=3)
    C <- A %*% cbind(1:3) + rnorm(4) # Adding a vector of length 4!
    ```
    
    Compute
    
    1.  Compute the matrix product between `A` and `B`
    2.  Generate the transpose of `A`
    3.  Merge `A` and `B` by row and by column. 
    3.  Calculate the element-wise product of `A` and the transpose of `B`.
    4.  Calculate The inverse of the cross product of `A`, this is $(A^\mbox{t}\times A)^{-1}$
    5.  Compute $(A^\mbox{t} A)^{-1} A^\mbox{t} C$. How could you do this with a single command in R?
  

    ```{r, eval=FALSE, echo=FALSE}
    # Matrix multiplication:
    A %*% B

    # Transpose
    t(A)
    
    # Element-wise product
    A*t(B)

    # Inverse
    solve(t(A) %*% A)
    
    # OLS
    solve(t(A) %*% A) %*% t(A) %*% C
    ```

# Matrix obects in R

Some language-specifics

- Finding the maximum value of every row: `max.col`

- Using the `which` function in matrices (set `arr.ind = TRUE`)

- Aggregation functions: `colSums`, `colMeans`, `row...`

- Other functions: `norm`, `lower.tri`

# Example: indexing columns

*   Imagine that you have a dataset with 100 rows and 5 columns (usually these
    numbers are bigger...)
    
*   Each cell reports the p-value of some association tests of the i-th row with
    the k-th column.
    
*   Task: For each row, select the column with the lowest value.
    
    ```{r eg-indexing-columns-dgp1}
    set.seed(11122)
    
    dat <- matrix(runif(100*5), nrow = 100)
    head(dat)
    ```
    
    ```{r eg-indexing-columns-sol1, include=FALSE}
    # To get the min is equivalent to get the max of the negative!
    idx <- max.col(-dat)
    head(idx)
    
    # Let's just see the first observations
    cbind(
      head(dat),
      lowest = head(dat[cbind(1:100, idx)])
    )
    ```


# Example: indexing rows

*   Now suppose that the data is a bit different: 5 rows and 100 columns.

*   Task: For each column, select the row with the highest value.
    
    ```{r eg-indexing-columns-dgp2}
    set.seed(881)
    
    dat <- matrix(runif(100*5), nrow = 5)
    str(dat)
    ```
    
    ```{r eg-indexing-columns-sol2, include = FALSE}
    # Well... we don't have a max.row... but we can always transpose it!
    idx <- max.col(-t(dat))
    head(idx)
    
    cbind(
      head(dat),
      lowest = head(dat[cbind(idx, 1:100)])
    )
    ```


# Problem 1

*   A bernoulli graph, also known as an Erdos-Renyi graph, is a simple model which
    is fully characterized by the density of the graph.

*   In a graph with probability parameter $p$ and $n$ nodes (vertices), the
    expected number of ties is $pn^2$, e.g. if $n = 100$ and $p = 0.1$
    $E(m) = 100^2 \times 0.1 = 1,000$.

*   Simulate a bernoulli network (a.k.a. Erdos-Renyi graph) with parameter p = .01
   (chance of any edge) and n = 1e3.
    
    ```{r solution-er, eval=FALSE, echo=FALSE}
    set.seed(81)
    # How many edges
    n   <- 1e3
    m   <- rbinom(1, n^2, .05)
    
    # Where are the edges?
    idx <- sample.int(n^2, m, FALSE)
    mat <- matrix(0, 1e3, 1e3)
    mat[idx] <- 1L
    
    image(mat)
    ```

# Example with a big graph

Let's imagine we want to do the same with a network of size 50,000 with density
.0005. The expected number of ties is 1,250,000.

```{r solution-er2}
set.seed(81)
# How many edges
n   <- 5e4
m   <- rbinom(1, n^2, .0005)

# Where are the edges?
idx <- sample.int(n^2, m, FALSE)

library(Matrix)
# Getting the row
idx_row <- (idx - 1) %% n + 1
idx_col <- (idx - 1) %/% n + 1

# Empty sparse
mat <- sparseMatrix(
  i = idx_row,
  j = idx_col,
  x = rep(1, length(idx_row)),
  dims = c(n, n)
  )

# Is it working?
sum(mat) == m
```

```{r er-space-saver}
print(object.size(mat), units = "Mb") # Sparse matrix
(8 * n^2) / 1024^2 # Dense matrix
```



# Questions 5

1.  Create a function to simulate OLS (ordinary least squares) data with the following:
    
    *   An argument `n`: Number of observations to make
    *   An argument `beta`: Coefficient with default value 0.3
    *   An argument `alpha`: Coefficient with default value 4
    *   `X` and the error are N(0, 1)
    *   The function should return both `Y` and `X`
    
    To generate 100 OLS model observations, you can type in
        
    ```r
    X <- rnorm(100)
    Y <- 4 + X*0.3 + rnorm(100)
    ```
    
2.  To make sure you are doing OK, try estimating your model using the `lm`
    function with a large simulation of observations.
        


# Thank you!

<div align="center">
R Bootcamp day 2<br>Statistical Computing Track<br><img src="../fig/trojan-rlogo.svg" style="width:250px;"><br>
</div>


<div align="center" style="font-size:40px;">
**website**: [ggvy.cl](https://ggvy.cl)<br><br>
**twitter**: [\@gvegayon](https://twitter.com/gvegayon)<br><br>
**github**: [gvegayon](https://github.com/gvegayon)<br><br>
</div>


# Session info

```{r}
devtools::session_info()
```

