---
title: "R Notebook"
output: html_notebook
---

This is an R Markdown Notebook. When you execute code within the notebook, the results appear beneath the code. 

Press the green button to install the packages we'll be using this week.

```{r}
install.packages(c("rmarkdown", "tidyverse", "sf", "ggfortify", "visdat", 
                   "plotly", "GGally", "janitor", "datasauRus", "gapminder", 
                   "nycflights13", "HistData", "devtools"))

# these packages lives on GitHub, not CRAN! Usually this just means they are in 
# development and will eventually be submitted to CRAN
devtools::install_github('thomasp85/gganimate') 
devtools::install_github('thomasp85/patchwork') 
```

The results of a code chunk will appear below. You can remove them by hitting the x button on the right. Since this is just installation stuff, go ahead.

Let's try code that produces something.

```{r}
plot(cars)
```


You can open a new R Notebook by going to **File > New File > R Notebook**. An R Notebook is a kind of R Markdown document. The only difference from a regular R Markdown file is the Preview button above. This will render an HTML document showing whatever was created by running the chunks most recently.

# Adding chunks

To add a new chunk, press *Cmd+Option+I* (*Ctrl+Alt+I* on Windows), or click the *Insert* button at the top of this document, then select *R*. 

Try making a code chunk below:



Nice! We'll use R Markdown for now, so you should put all your code in chunks.

# HTML version

When you save the notebook, an HTML file containing the code and output will be saved alongside it. This makes a nice, polished report of your work to share.

Click the *Preview* button at the top of this document or press *Cmd+Shift+K* (*Ctrl+Shift+K* on Windows) to preview the HTML file. Note this will only work _after_ you save exercises.Rmd (because the act of saving generates/updates the file to preview. Try saving and clicking *Preview* now.

