---
title: "R Notebook"
output: html_notebook
---

This is an R Markdown Notebook. When you execute code within the notebook, the results appear beneath the code. 

Press the green button to install the packages we'll be using.

```{r}
install.packages(c("rmarkdown", "tidyverse", "sf", "datasauRus", "gapminder", "nycflights13", "devtools"))
devtools::install_github('thomasp85/gganimate') # this package lives on GitHub, not CRAN!
```

The results of a code chunk will appear below. You can remove them by hitting the x button on the right. Since this is just installation stuff, go ahead.

Let's try code that produces something.

```{r}
plot(cars)
```


You can open a new R Notebook by going to **File > New File > R Notebook**. An R Notebook is a kind of R Markdown document. The only difference from a regular R Markdown file is the Preview button above. This will render an HTML document showing whatever was created by running the chunks most recently.

# Adding chunks

To add a new chunk, press *Cmd+Option+I* (*Ctrl+Alt+I* on Windows), or click the *Insert* button at the top of this document, then select *R*. 

Try making a code chunk below:



Nice! We'll use R Markdown for now, so you should put all your code in chunks.

# HTML version

When you save the notebook, an HTML file containing the code and output will be saved alongside it. This makes a nice, polished report of your work to share.

Click the *Preview* button at the top of this document or press *Cmd+Shift+K* (*Ctrl+Shift+K* on Windows) to preview the HTML file. Note this will only work _after_ you save exercises.Rmd (because the act of saving generates/updates the file to preview. Try saving and clicking *Preview* now.

# Packages

You can immediately run any function from base R within a notebook, But if you'd like to run a function that comes in an R package, you will need to first load the package in the notebook.

Do you remember how to run the core tidyverse packages? Load them in the chunk below:

```{r}
# run the tidyverse package


```

In R, you need to reload your packages every time you start a new session. It's a good habit to start new sessions every time so that you know your code is working without bugs!

# R Basics

in R, you can save the results of just about any function. R let's you use `=` or `<-` for assignment to a variable name. Our preference is for `<-`, but it's more important to be consistent than which you use. 

You can use *Option+-* (*Alt+-* on Windows) to insert the assignment arrow, `<-`.

```{r}
cars2 = cars
cars3 <- cars
identical(cars2, cars3) # are these two identical? TRUE or FALSE
```

Use `?` in front of a function name to see the documentation. You can also highlight the function name and press F1.

```{r}
?plot
```

Function arguments use `=`. You don't have to specify a name if the arguments are in order, but it's better to do so after the second or third argument you have in your function.

```{r}
mean(cars3$speed, na.rm = TRUE)
```

