---
title: Day 2
author: George G. Vega Yon
output:
  slidy_presentation:
    footer: IMAGE
    highlight: pygments
---

We will work on simulations!

Today's goals are:

1.  Learn more about base R, in particular
1.  Use extarnal data: `read.table`,
2.  Data structures: `vector`, `matrix`, `list`
3.  Functional programming: Creating functions in R,
2.  Control flow: `for` loops (we'll leave ),
3.  Pseudo random numbers: `sample`, `runif`, `rnorm`
4.  Summarizing: `boxplot`

5.  Extra 1: Applying parallel computing
6.  Extra 6: Animating our simulation with `magick`.


We'll do all of this by runing some monte carlo simulations models...

# Statistical Functions

*   R has a very nice set of families of distributions. In general, distribution functions have the following name structure:
    
    a.  **Random Number Generation**: `r[name-of-the-distribution]`, e.g. `rnorm` for normal, `runif` for uniform.
    
    b.  **Density function**: `d[name-of-the-distribution]`, e.g. `dnorm` for normal, `dunif` for uniform.
    
    c.  **Cumulative Distribution Function (CDF)**: `p[name-of-the-distribution]`, e.g. `pnorm` for normal, `punif` for uniform.
    
    d.  **Inverse of the CDF (quantile)**: `q[name-of-the-distribution]`, e.g. `qnorm` for the normal, `qunif` for the uniform.
    
*   In the case of [**pseudo random numbers**](https://en.wikipedia.org/wiki/Pseudorandom_number_generator){target="_blank"} (the `r` prefix), it is important to **always** set the seed to ensure **reproducibility**

    ```{r setseed, collapse=TRUE}
    # First run
    set.seed(12)
    rnorm(4)
    
    # Second run
    set.seed(12)
    rnorm(4)
    ```
    
    More distributions available at `??Distributions`.

---

*   Take a look at the normal distribution

```{r}
set.seed(12)
op <- par(mfrow = c(2,2))
hist(rnorm(1e5))
curve(qnorm)
curve(pnorm, xlim=c(-3, 3))
curve(dnorm, xlim=c(-3, 3))
par(op)
```

---

*   Take a look at the exponential

```{r}
set.seed(12)
op <- par(mfrow = c(2,2))
hist(rexp(1e5))
curve(qexp)
curve(pexp, xlim=c(0, 6))
curve(dexp, xlim=c(0, 6))
par(op)
```

# Questions

1.  Draw 1e5 samples from a chi2 with 2 degrees of freedom (hint: check `?Distributions`).

```{r, echo=FALSE, eval=FALSE}
set.seed(778)
Z <- rchisq(1e5, 2)
hist(Z, breaks=100)
```
    

2.  Draw 1e5 samples from a chi2 with 2 degrees of freedom using `rnorm` (hint: Recall that if $X\sim N(0,1)$, then $X^2\sim\chi^2_1$, and if $X, Y\sim N(0,1)$, then $X^2 + Y^2\sim\chi^2_2$).

```{r, echo=FALSE, eval=FALSE}
set.seed(778)
Z <- rnorm(1e5)^2 + rnorm(1e5)^2
hist(Z, breaks=100)
```

# Retirement Savings

![Courtesy of [whatacatchjohnny-blog](http://whatacatchjohnny-blog.tumblr.com/post/7104244664/the-devils-advocate)](free-will.gif){.class width="600px"}

*   The hyperbolic discount problem: In simple words, people overvalue the present (which yields regret in the future!)

*   Online tools for projecting retirement savings are usually very optimistic (sorry!), like yielding a 5\% fix average return (no bond gives you that!)

*   We will perform some simulation models using real data and, instead of obtaining a point estimate of our future savings, we will obtain confidence intervals (a range of possible outcomes).


# Today's tasks

1.  Read in the data from returns from the stocks market.

2.  Create a function that simulates a random return trejectory by sampling from the stocks market.

3.  Mix all in a for-loop function that creates a random pension savings trajectory.

4.  Wrap it around 1,000 simulation, and plot the output using the `boxplot` function in R.


# About the dataset

*   Dow Jones Industrial Average (downloaded from https://quotes.wsj.com/index/DJIA/historical-prices)
    
    -   This dataset covers from 1990 to 2018
    -   It provides the prices at the daily
    
    ```{r}
    # Loading the R package
    library(readr)
    
    # Reading the data
    dow <- read_csv(
      "HistoricalPrices.csv",
      # We can tell the function how the data is stored
      col_types = cols(Date = col_date(format = "%m/%d/%y"))
      )
    
    # Let's print it out 
    dow
    
    # Taking a look at it
    plot(x = dow$Date, y = dow$Close, type="l")
    ```
    
# Keeping the monthly data only

```{r}
library(magrittr)
library(dplyr)
dow_monthly <- dow %>%
  filter(months(Date) != months(lag(Date))) %>% # Keeping the last day of the month
  arrange(Date) %>%                             # Sorting by date
  mutate(
    r = Close/lag(Close) - 1                    # Computing the monthly return
  ) %$%
  r[-1]                                         # Keeping all, but the last month   
```

# About the model

$$
S_t = S_{t-1}\times(1 + r_t) + I_t\times c_t
$$

Where $S_t$ is your savings balance in time $t$, $r_t\sim R$ is the return rate in time $t$, $I_t$ is your income in time $t$, and $c_t$ is your contribution rate in time $t$.

# Functional Programming

```{r dow-function}
sample_dow <- function(n=100) {
  sample(dow_monthly, n, replace = TRUE)
}
```

Creating the core function


```{r savings-sim} 
#' @param n_nomths Integer, number of months to simulate.
#' @param S0 Initial balance of savings.
#' @param I0 Income.
#' @param c_rate Monthly contribution rate
sim_savings <- function(
  n_months = 200,
  S0       = 0,
  I0       = 40/12,
  c_rate   = .05
) {
  
  # Generating returns
  R <- sample_dow(n_months)
  S <- vector("numeric", n_months)
  
  # First value
  S[1] <- S0
  
  for (i in 2:n_months) 
    S[i] <- S[i - 1]*(1 + R[i - 1]) + I0*c_rate
  
  return(S)
  
}

plot(sim_savings(200), type="l")
```

# Simulations

Running multiple simulations using the `replicate` function.

```{r multiple-simulations}
set.seed(1)
nsim    <- 1e4
nmonths <- 12*30

ans02 <- replicate(nsim, sim_savings(nmonths, c_rate = 0.025))
ans05 <- replicate(nsim, sim_savings(nmonths, c_rate = 0.05))
ans10 <- replicate(nsim, sim_savings(nmonths, c_rate = 0.10))
ans15 <- replicate(nsim, sim_savings(nmonths, c_rate = 0.15))

```

# Analyzing the results

Stacking the last row (period) of each set of simulations using `cbind` 

```{r stacking-the-results}
ans <- cbind(
  "2.5%" = ans02[nmonths,],
  "5%"   = ans05[nmonths,],
  "10%"  = ans10[nmonths,],
  "15%"  = ans15[nmonths,]
  )

# We can look at a few observations with the `head` function
head(ans)
```


# Visualizing the output

```{r}
plot_CI <- function(x, conf_interval=.95, col="steelblue")  {
  
  # Computing cureves
  lvl <- (1-conf_interval)/2
  lvl <- c(lvl, .5, 1-lvl)
  dat <- t(apply(x, 2, quantile, lvl))
  
  # Emptyplot
  plot(NA,xlim = c(1,nrow(dat)), ylim=c(0, max(dat)))
  
  # Drawing the confidence region
  polygon(
    c(1:nrow(dat), nrow(dat):1),
    c(dat[,1], rev(dat[,3])),
    col    = adjustcolor(col, alpha.f = .5),
    border = "transparent"
  )
  
  # Drawing the 50%
  lines(dat[,2], lwd=2, lty=2, col=col)
  
  # Legend
  legend(
    "topleft",
    col  = c(col, col),
    lty  = c(NA, 2),
    fill = c(col, NA),
    legend = c(
      paste0("%", conf_interval*100, " confidence"),
      "Median"
      )
  )
}

op <- par(mfrow=c(2,2))
plot_CI(t(ans02))
plot_CI(t(ans05))
plot_CI(t(ans10))
plot_CI(t(ans15))
par(op)
```


```{r plotting-the-outcome}
# Changing the palette
palette(
  adjustcolor(RColorBrewer::brewer.pal(8, "Set1"), alpha=.7)
  )

# Using a nice-looking boxplot
boxplot(
  ans,
  main = "Distribution of Retirement Savings Balance",
  xlab = "Monthly Savings as a Proportion of Income",
  ylab = "Savings (thousands of USD, log-scale)",
  log  = "y",
  las  = 2,
  col  = 1:4,
  lwd  = 1.5,
  border = 1:4
)
```

    

# Problems

1.  In a lot of cases, more than encouraging people to increase their savings is
to ask them to retire later. Using the code shown here, analyze what are the effects
of postponing retirement 1, 3, and 5 years.

2.  A big assumption we made was having no gaps, this is, always employed and
contributing. To account for this, we can include a poisson process as follows:
    a.  Simulate times at which you become unemployed assuming that on average,
    you will spend half a month unemployed, i.e. a rate of (1/24).
    b.  You can do this using the `rexp` function, e.g.:
    ```{r poisson-process}
    set.seed(1231)
    when_unemployed <- rexp(5, 1/24) # Times at which become unemployed
    when_unemployed <- cumsum(floor(when_unemployed))
    when_unemployed 
    ```
    
    Assume that you spend one month unemployed.

3.  
    
    
