---
title: Day 2
author: George G. Vega Yon
---

We will work on simulations!

Today's goals are:

1.  Learn more about base R, in particular
1.  Use extarnal data: `read.table`,
2.  Data structures: `vector`, `matrix`, `list`
3.  Functional programming: Creating functions in R,
2.  Control flow: `for` loops (we'll leave ),
3.  Pseudo random numbers: `sample`, `runif`, `rnorm`
4.  Summarizing: `boxplot`

5.  Extra 1: Applying parallel computing
6.  Extra 6: Animating our simulation with `magick`.


We'll do all of this by runing some monte carlo simulations models...

# Retirement Savings

![Courtesy of [whatacatchjohnny-blog](http://whatacatchjohnny-blog.tumblr.com/post/7104244664/the-devils-advocate)](free-will.gif){.class width="600px"}

*   The hyperbolic discount problem: In simple words, people overvalue the present (which yields regret in the future!)

*   Online tools for projecting retirement savings are usually very optimistic (sorry!), like yielding a 5\% fix average return (no bond gives you that!)

*   We will perform some simulation models using real data and, instead of obtaining a point estimate of our future savings, we will obtain confidence intervals (a range of possible outcomes).


# Today's tasks

1.  Read in the data from returns from the stocks market.

2.  Create a function that simulates a random return trejectory by sampling from the stocks market.

3.  Mix all in a for-loop function that creates a random pension savings trajectory.

4.  Wrap it around 1,000 simulation, and plot the output using the `boxplot` function in R.


# About the dataset

*   Dow Jones Industrial Average (downloaded from https://quotes.wsj.com/index/DJIA/historical-prices)
    
    -   This dataset covers from 1990 to 2018
    -   It provides the prices at the daily
    
    ```{r}
    # Loading the R package
    library(readr)
    
    # Reading the data
    dow <- read_csv(
      "HistoricalPrices.csv",
      # We can tell the function how the data is stored
      col_types = cols(Date = col_date(format = "%m/%d/%y"))
      )
    
    # Let's print it out 
    dow
    
    # Taking a look at it
    plot(x = dow$Date, y = dow$Close, type="l")
    ```
    
# Keeping the monthly data only

```{r}
library(magrittr)
library(dplyr)
dow_monthly <- dow %>%
  filter(months(Date) != months(lag(Date))) %>%
  arrange(Date) %>%
  mutate(
    r = Close/lag(Close) - 1
  ) %$%
  r[-1]
```

# About the model

$$
S_t = S_{t-1}\times(1 + r_t) + I_t\times c_t
$$

Where $S_t$ is your savings balance in time $t$, $r_t\sim R$ is the return rate in time $t$, $I_t$ is your income in time $t$, and $c_t$ is your contribution rate in time $t$.

# Functional Programming

```{r dow-function}
sample_dow <- function(n=100) {
  sample(dow_monthly, n, replace = TRUE)
}
```

Creating the core function


```{r savings-sim} 
#' @param n_nomths Integer, number of months to simulate.
#' @param S0 Initial balance of savings.
#' @param I0 Income.
#' @param c_rate Monthly contribution rate
sim_savings <- function(
  n_months = 200,
  S0       = 0,
  I0       = 40/12,
  c_rate   = .05
) {
  
  # Generating returns
  R <- sample_dow(n_months)
  S <- vector("numeric", n_months)
  
  # First value
  S[1] <- S0
  
  for (i in 2:n_months) 
    S[i] <- S[i - 1]*(1 + R[i - 1]) + I0*c_rate
  
  return(S)
  
}

plot(sim_savings(200), type="l")
```

# Simulations

Running multiple simulations using the `replicate` function.

```{r multiple-simulations}
set.seed(1)
nsim    <- 1e4
nmonths <- 12*30

ans02 <- replicate(nsim, sim_savings(nmonths, c_rate = 0.025))
ans05 <- replicate(nsim, sim_savings(nmonths, c_rate = 0.05))
ans10 <- replicate(nsim, sim_savings(nmonths, c_rate = 0.10))
ans15 <- replicate(nsim, sim_savings(nmonths, c_rate = 0.15))

```

# Analyzing the results

Stacking the last row (period) of each set of simulations using `cbind` 

```{r stacking-the-results}
ans <- cbind(
  "2.5%" = ans02[nmonths,],
  "5%"   = ans05[nmonths,],
  "10%"  = ans10[nmonths,],
  "15%"  = ans15[nmonths,]
  )

# We can look at a few observations with the `head` function
head(ans)
```


# Visualizing the output

```{r plotting-the-outcome}
# Changing the palette
palette(
  adjustcolor(RColorBrewer::brewer.pal(8, "Set1"), alpha=.7)
  )

# Using a nice-looking boxplot
boxplot(
  ans,
  main = "Distribution of Retirement Savings Balance",
  xlab = "Monthly Savings as a Proportion of Income",
  ylab = "Savings (thousands of USD, log-scale)",
  log = "y",
  las = 2,
  col = 1:4,
  lwd = 1.5,
  border = 1:4
)
```

    

