---
title: "Analyze the Effect of Right Heart Catherization on Survival"
output: html_document
keywords: [analysis]
difficulty: medium
author: Malcolm Barrett
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# load any packages you need
library(tidyverse)
```

## Does Right Heart Catherization Improve Survival?

Included in this folder is a data set, `rhc`, which contains data related to survival after receiving a treatment called right heart catherization (RHC).

You can find a data dictionary on the [Vanderbilt website](http://biostat.mc.vanderbilt.edu/wiki/pub/Main/DataSets/Crhc.html)

From the data description: 

> "This dataset was used in Connors et al. (1996): The effectiveness of RHC in the initial care of critically ill patients. The dataset pertains to day 1 of hospitalization, i.e., the 'treatment' variable `swang1` is whether or not a patient received a RHC (also called the Swan-Ganz catheter) on the first day in which the patient qualified for the SUPPORT study."

You can consider this a survival analysis (using Cox regression) or a simple binary outcome (using logistic regression). For a simple logistic regression, use the outcome variable `survived`. If you choose to do a Cox regression, use the survival package. The model will look like:

```{r, eval = FALSE}
coxph(Surv(t3d30, dth30) ~ swang1 + <COVARIATES>, data = rhc)
```

Otherwise, just use `glm()`.

## Analysis

1. Import the data.
2. Explore the data.
3. Fit a model for the effect of `swang1` (RHC) on survival. Include covariates you think might be important. It's ok if you don't know much about cardiology; just use your instinct for what's important.
4. Present the results of the model in a table or plot.