---
title: "Transform Data"
output:
  html_document:
    df_print: paged
---

```{r setup}
library(tidyverse)
library(gapminder)
library(nycflights13)

# Toy datasets to use

superheroes <- tribble(
       ~name, ~alignment,  ~gender,          ~publisher,
   "Magneto",      "bad",   "male",            "Marvel",
     "Storm",     "good", "female",            "Marvel",
  "Mystique",      "bad", "female",            "Marvel",
    "Batman",     "good",   "male",                "DC",
     "Joker",      "bad",   "male",                "DC",
  "Catwoman",      "bad", "female",                "DC",
   "Hellboy",     "good",   "male", "Dark Horse Comics"
  )

publishers <- tribble(
  ~publisher, ~yr_founded,
        "DC",       1934L,
    "Marvel",       1939L,
     "Image",       1992L
  )

```

## gapminder

```{r}
gapminder
```


## Your Turn 1

Alter the code to select just the `pop` column:

```{r}
select(gapminder, year, lifeExp)
```

## Quiz

Which of these is NOT a way to select the `name` and `n` columns together?

```{r}
select(gapminder, -c(year, lifeExp, pop, gdpPercap))
select(gapminder, country:continent)
select(gapminder, starts_with("c"))
select(gapminder, ends_with("t"))
```

## Your Turn 2

Show:

* All of the rows where `pop` is greater than or equal to 100000  
* All of the rows for El Salvador  
* All of the rows that have a missing value for `year`  

```{r}


filter(gapminder, is.na(year))
```

## Your Turn 3

Use Boolean operators to alter the code below to return only the rows that contain:

* El Salvador  
* Countries that had populations over 100000 in 1960 or earlier
* Countries that are one of El Salvador, Oman, or Croatia

```{r}
filter(gapminder, country == "El Salvador" | country == "Oman")


```

## Your Turn 4

Arrange gapminder by `year`. Add `lifeExp` as a second (tie breaking) variable to arrange on. Which country had the lowest life expectency in 1952?

```{r}

```

## Your Turn 5

Use `desc()` to find the country with the highest `gdpPercap`.
Then, use `desc()` to the country with the highest `lifeExp`

```{r}

```

## Your Turn 6

Use `%>%` to write a sequence of functions that: 

1. Filter only countries thate are in the continent of Oceania.
2. Select the `country`, `year` and `lifeExp` columns  
3. Arrange the results so that the highest life expetency is at the top.

```{r}

```

## Exam

1. Import data from...

```{r}

```

## Your Turn 7

Use summarise() to compute three statistics about the data:

1. The first (`min()`) year in the data
2. The last (`max()`) year in the data  
3. The total number of observations (`n()`) and the total number of unique countries in the data (`n_distinct()`)

```{r}

```

## Your Turn 8

Extract the rows where `continent == "Europe"`. Then use `group_by()` to group by country. Finally, use `summarize()` to compute:

1. The total number of observations for each country in Europe
2. The lowest observed life expectency for each country 

```{r}

```

## Your Turn 9

Use `group_by()`, `summarize()`, and `arrange()` to display the continents with the highest average life expectency.

```{r}

```

## Your Turn 10

Use grouping to calculate the mean life expectancy for each year by continent. Call the mean life expectancy variable `mean_le`. Plot the life expectancy over time (no need to change the plot code).

```{r}
gapminder %>% 
  __________ %>% 
  __________ %>% 
  ggplot(aes(x = year, y = mean_le, col = continent)) +
    geom_line() + 
    scale_color_manual(values = continent_colors)
```

## Your Turn 11

Which airlines had the largest arrival delays? Complete the code below.

1. Join `airlines` to `flights`
2. Compute and order the average arrival delays by airline. Display full names, no codes.

```{r}
flights %>%
  drop_na(arr_delay) %>%
                      %>%
  group_by(         ) %>%
                      %>%
  arrange(       ) 
```

## Your Turn 12

How many airports in `airports` are serviced by flights originating in New York (i.e. flights in our dataset?) Notice that the column to join on is named `faa` in the **airports** data set and `dest` in the **flights** data set.


```{r}
__________ %>%
 _________(_________, by = ___________) %>%
  distinct(faa)
```



***

# Take aways

* Extract variables with `select()`  
* Extract rows with `filter()`  
* Arrange rows, with `arrange()`  

* Make tables of summaries with `summarise()`  
* Make new variables, with `mutate()`  
* Do groupwise operations with `group_by()`

* Connect operations with `%>%`  

* Use `left_join()`, `right_join()`, `full_join()`, or `inner_join()` to join datasets
* Use `semi_join()` or `anti_join()` to filter datasets against each other


